---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# punycoder

<!-- badges: start -->
[![Verify](https://github.com/bart-turczynski/punycoder/actions/workflows/verify.yml/badge.svg)](https://github.com/bart-turczynski/punycoder/actions/workflows/verify.yml)
[![CRAN status](https://www.r-pkg.org/badges/version/punycoder)](https://CRAN.R-project.org/package=punycoder)
<!-- badges: end -->

High-performance Unicode and Punycode encoding/decoding for internationalized domain names (IDNs) in R.

## Overview

The `punycoder` package addresses critical gaps in R's URL processing capabilities by providing reliable, fast conversion between Unicode and ASCII representations of domain names. It follows RFC 3492 standards and is designed for robust handling of internationalized domain names in web scraping, data analysis, and URL processing workflows.

## Dependencies

`punycoder` has a small dependency footprint:

- Runtime dependencies: `R (>= 3.5.0)`, `Rcpp`
- Optional system dependency: `libidn2` (detected at compile time)
- Optional build helper: `pkg-config` (used by `configure` to detect `libidn2`)
- Development dependencies: `testthat`, `knitr`, `rmarkdown`

## Installation

Install the released version of punycoder from
[CRAN](https://CRAN.R-project.org/package=punycoder) with:

```{r eval=FALSE}
install.packages("punycoder")
```

Or install the development version from
[GitHub](https://github.com/bart-turczynski/punycoder) with:

```{r eval=FALSE}
# install.packages("remotes")
remotes::install_github("bart-turczynski/punycoder")
```

### Optional native backend (`libidn2`)

`punycoder` works without extra system libraries. If `libidn2` is available at
build time, the package enables a native backend automatically; otherwise it
uses the built-in C++ fallback backend.

To install the recommended optional dependency:

- macOS (Homebrew):
  - `brew install libidn2 pkg-config`
- Debian/Ubuntu:
  - `sudo apt-get install libidn2-0-dev pkg-config`
- Fedora/RHEL/CentOS:
  - `sudo dnf install libidn2-devel pkgconf-pkg-config`
- Arch Linux:
  - `sudo pacman -S libidn2 pkgconf`

Verify the library is visible before installing `punycoder` from source:

```{r eval=FALSE}
system("pkg-config --modversion libidn2")
```

Then install/reinstall `punycoder`:

```{r eval=FALSE}
remotes::install_github("bart-turczynski/punycoder")
```

## Example

```{r example}
library(punycoder)

# Basic encoding
puny_encode("café.com")

# Check if domain is punycode
is_punycode("xn--example")

# Validate domains
validate_domain("test.com")
```

## Key Features

- **Reliable Encoding/Decoding**: RFC 3492 compliant punycode conversion
- **Best-effort host rewriting**: Swap the host of a URL-shaped string in place (not a full URL parser; see below)
- **High Performance**: Vectorized operations for processing large datasets
- **Comprehensive Validation**: Robust error handling with informative messages
- **Flexible Backend**: Automatically uses `libidn2` when available, with a built-in fallback backend

## Use Cases

### Web Scraping
Process international websites with Unicode domain names:

```{r eval=FALSE}
international_urls <- c(
  "https://café.paris.fr/menu",
  "https://москва.рф/news",
  "https://北京.中国/info"
)

# Convert for HTTP requests (best-effort host rewriting only)
ascii_urls <- url_encode(international_urls)
```

> `url_encode()`, `url_decode()`, and `parse_url()` do **best-effort host
> extraction and rewriting**, not RFC 3986 / WHATWG URL parsing or
> canonicalization. They have no percent encoding/decoding, scheme validation,
> robust port/path/query semantics, full IPv6 (zone IDs / RFC 6874), or
> serialization guarantees, and are slated for eventual removal in favour of a
> dedicated URL package consuming punycoder's host functions. Use
> `host_normalize()` / `puny_encode()` directly when you control the host
> parse.

### Data Analysis
Clean and standardize URL datasets:

```{r eval=FALSE}
# Identify international domains
is_idn(c("café.com", "example.com", "москва.рф"))

# Validate domain names
validate_domain(c("valid.com", "invalid..domain"))
```

## Current State

`punycoder` currently provides:

- Domain encoding/decoding: `puny_encode()`, `puny_decode()`
- Best-effort URL host rewriting/extraction (not URL parsing/canonicalization): `url_encode()`, `url_decode()`, `parse_url()`
- Domain validation utilities: `is_punycode()`, `is_idn()`, `validate_domain()`
- Vectorized operations and strict/non-strict handling for malformed input
- Build-time backend selection (`libidn2` when present, built-in fallback otherwise)
- Best-effort structured host extraction where invalid inputs are returned as missing components

## Non-goals

`punycoder` is a standards primitive for Punycode and host normalization. It is
deliberately agnostic about resolvability and safety; the following are **not**
part of its acceptance criteria:

- **No spoof / homograph / mixed-script / display-safety detection.**
  `host_normalize()` is not a safety gate — a successful result says the host is
  valid and normalized under the pinned UTS #46 profile, nothing about whether it
  is visually safe or non-deceptive. Confusable and restriction-level checks
  (UTS #39 / UTR #36, which UTS #46 itself recommends only as application/UI-layer
  steps) belong upstack.
- **No URL canonicalization.** The `url_*` / `parse_url()` helpers do best-effort
  host rewriting only (see above), not RFC 3986 / WHATWG URL parsing.
- **No DNS resolvability or registrability / PSL classification.**

These opinions belong in higher layers that consume punycoder's host functions.

## Acknowledgments

- Core C++/R integration is powered by `Rcpp`.
- Optional native punycode backend support is provided through `libidn2`.
- `punycoder` is inspired by `urltools` and is designed to provide a robust
  fix for punycode encode/decode issues that may arise in `urltools`
  workflows.

## Contributing

We welcome contributions. See [CONTRIBUTING.md](https://github.com/bart-turczynski/punycoder/blob/main/CONTRIBUTING.md) for the
current development workflow.

## License

MIT 
