---
output:
  github_document:
    pandoc_args: ["--wrap=preserve"]
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# punycoder

<!-- badges: start -->
[![CRAN status](https://www.r-pkg.org/badges/version/punycoder)](https://CRAN.R-project.org/package=punycoder)
[![CRAN downloads](https://cranlogs.r-pkg.org/badges/punycoder)](https://CRAN.R-project.org/package=punycoder)
[![CRAN checks](https://badges.cranchecks.info/worst/punycoder.svg)](https://cran.r-project.org/web/checks/check_results_punycoder.html)
[![r-universe](https://bart-turczynski.r-universe.dev/punycoder/badges/version)](https://bart-turczynski.r-universe.dev/punycoder)
[![Pipeline](https://gitlab.com/bart-turczynski/punycoder/badges/main/pipeline.svg)](https://gitlab.com/bart-turczynski/punycoder/-/pipelines)
[![Coverage](https://gitlab.com/bart-turczynski/punycoder/badges/main/coverage.svg)](https://gitlab.com/bart-turczynski/punycoder/-/pipelines)
[![Docs](https://img.shields.io/website?url=https%3A%2F%2Fbart-turczynski.gitlab.io%2Fpunycoder%2F&label=docs&logo=gitlab&logoColor=white&up_message=pkgdown&up_color=1f75cb)](https://bart-turczynski.gitlab.io/punycoder/)
[![Latest release](https://img.shields.io/gitlab/v/release/bart-turczynski%2Fpunycoder)](https://gitlab.com/bart-turczynski/punycoder/-/releases)
[![Lifecycle: stable](https://img.shields.io/badge/lifecycle-stable-brightgreen.svg)](https://lifecycle.r-lib.org/articles/stages.html#stable)
[![Project Status: Active](https://www.repostatus.org/badges/latest/active.svg)](https://www.repostatus.org/#active)
[![DOI](https://zenodo.org/badge/DOI/10.5281/zenodo.20973629.svg)](https://doi.org/10.5281/zenodo.20973629)
[![Zenodo](https://img.shields.io/badge/Zenodo-all_software-1682D4?logo=zenodo&logoColor=white)](https://zenodo.org/search?q=metadata.creators.person_or_org.identifiers.identifier:0000-0002-8788-7980)
[![OpenSSF Best Practices](https://www.bestpractices.dev/projects/13429/badge)](https://www.bestpractices.dev/projects/13429)
[![License](https://img.shields.io/gitlab/license/bart-turczynski%2Fpunycoder)](https://gitlab.com/bart-turczynski/punycoder/-/blob/main/LICENSE.md)
[![Dependencies](https://tinyverse.netlify.app/badge/punycoder)](https://CRAN.R-project.org/package=punycoder)
[![Last commit](https://img.shields.io/gitlab/last-commit/bart-turczynski%2Fpunycoder)](https://gitlab.com/bart-turczynski/punycoder/-/commits/main)
<!-- badges: end -->

High-performance Unicode and Punycode encoding/decoding for internationalized domain names (IDNs) in R.

## Overview

The `punycoder` package provides fast, standards-based conversion between Unicode and ASCII representations of domain names, across two distinct surfaces:

- a **low-level Punycode codec** — `puny_encode()` / `puny_decode()` — the raw RFC 3492 transform with `xn--` A-label framing (RFC 5890/5891) and letter-digit-hyphen checks, **not** an IDNA normalization API (no Unicode NFC, UTS #46 mapping, or case folding);
- an **IDNA/UTS-46 host-normalization surface** — `host_normalize()` — mapping a host name to its canonical lowercase ASCII comparison form under a pinned UTS #46 non-transitional profile.

`host_normalize()` is a **UTS #46 profile, not IDNA2008 conformance** — UTS #46 is compatibility processing and deliberately accepts labels IDNA2008 would reject (e.g. `☕.example` → `xn--53h.example`). See `?host_normalize` and `normalization_profile_info()` for the normative profile and full standards references (RFC 3492/5890/5891/5892/5893, UTS #46, UAX #15/#44, STD 3, RFC 8753).

Normalization runs against vendored Unicode data, and a build ships a *set* of Unicode versions with one pinned as the default (currently 17.0.0, with 16.0.0 also shipped). `unicode_versions()` reports what the installed build carries, and `host_normalize(x, unicode_version = "16.0.0")` selects another one for a single call — which `normalization_profile_info()` reflects by appending `+unicode-<version>` to the reported profile token.

## Installation

Install the released version of punycoder from
[CRAN](https://CRAN.R-project.org/package=punycoder) with:

```{r eval=FALSE}
install.packages("punycoder")
```

Or install the development version from
[r-universe](https://bart-turczynski.r-universe.dev/punycoder) with:

```{r eval=FALSE}
install.packages(
  "punycoder",
  repos = c("https://bart-turczynski.r-universe.dev", "https://cloud.r-project.org")
)
```

To build it from source on
[GitLab](https://gitlab.com/bart-turczynski/punycoder) instead:

```{r eval=FALSE}
# install.packages("remotes")
remotes::install_gitlab("bart-turczynski/punycoder")
```

### System requirements

`punycoder` needs R 4.1.0 or later and `Rcpp`, which `install.packages()` pulls
in. A binary install needs nothing else. A source install needs a C++ compiler,
which the standard R build tools for each platform provide. `libidn2` 2.3.5 or
later and `pkg-config` are optional: `configure` uses them for a native backend
when it finds both.

### Optional native backend (`libidn2`)

`punycoder` works without extra system libraries. If `libidn2` is available at
build time, the package enables a native backend automatically; otherwise it
uses the built-in C++ fallback backend.

To install the recommended optional dependency:

- macOS (Homebrew):
  - `brew install libidn2 pkg-config`
- Debian/Ubuntu:
  - `sudo apt-get install libidn2-0-dev pkg-config`
- Fedora/RHEL/CentOS:
  - `sudo dnf install libidn2-devel pkgconf-pkg-config`
- Arch Linux:
  - `sudo pacman -S libidn2 pkgconf`

Verify the library is visible before installing `punycoder` from source:

```{r eval=FALSE}
system("pkg-config --modversion libidn2")
```

Then install/reinstall `punycoder`:

```{r eval=FALSE}
remotes::install_gitlab("bart-turczynski/punycoder")
```

## Example

```{r example}
library(punycoder)

# Basic encoding
puny_encode("café.com")

# Check if domain is punycode
is_punycode("xn--example")

# Validate domains
validate_domain("test.com")
```

### Command-line use

`punycoder` is a library, not a CLI, but a one-line `Rscript` wrapper covers the
usual shell need: turning an IDN into the ASCII form that `dig`, `nslookup`, or
`curl` will actually accept. No extra install, no `bin/` script. The snippets
below are POSIX `sh`/`bash` — keep the `-e` argument in single quotes so the
shell leaves the R code alone.

Unicode host → ASCII, via the UTS #46 canonical form:

```sh
Rscript -e 'cat(punycoder::host_normalize(commandArgs(TRUE)), sep = "\n")' münchen.de
#> xn--mnchen-3ya.de
```

Use `puny_encode()` instead when you want the raw RFC 3492 transform with no
UTS #46 mapping, and `puny_decode()` to go back the other way:

```sh
Rscript -e 'cat(punycoder::puny_encode(commandArgs(TRUE)), sep = "\n")' münchen.de
#> xn--mnchen-3ya.de

Rscript -e 'cat(punycoder::puny_decode(commandArgs(TRUE)), sep = "\n")' xn--mnchen-3ya.de
#> münchen.de
```

Feeding a resolver is then just command substitution (drop `sep` so nothing but
the host is printed):

```sh
dig +short "$(Rscript -e 'cat(punycoder::host_normalize(commandArgs(TRUE)))' münchen.de)"
#> 194.246.166.100

nslookup "$(Rscript -e 'cat(punycoder::host_normalize(commandArgs(TRUE)))' münchen.de)"
```

All three functions are vectorized, so several hosts can go through one R
startup and out to `xargs`:

```sh
Rscript -e 'cat(punycoder::host_normalize(commandArgs(TRUE)), sep = "\n")' münchen.de москва.рф |
  xargs -n 1 dig +short
```

#### Failed conversions

`host_normalize()` never aborts: an input it cannot normalize comes back as
`NA`, which prints as a literal `NA` line rather than silently disappearing from
the pipeline.

```sh
Rscript -e 'cat(punycoder::host_normalize(commandArgs(TRUE)), sep = "\n")' münchen.de 'bad..domain'
#> xn--mnchen-3ya.de
#> NA
```

`puny_encode()` and `puny_decode()` behave differently: they are strict by
default, so a bad input stops the script with a non-zero exit status.

```sh
Rscript -e 'cat(punycoder::puny_encode(commandArgs(TRUE)), sep = "\n")' 'bad..domain'
#> Error: Error encoding domain: Domain contains empty label
#> Execution halted
```

Pass `strict = FALSE` to get the per-element `NA` behavior instead, so one bad
host does not take down a batch:

```sh
Rscript -e 'cat(punycoder::puny_encode(commandArgs(TRUE), strict = FALSE), sep = "\n")' münchen.de 'bad..domain'
#> xn--mnchen-3ya.de
#> NA
```

Since `NA` is printed, not swallowed, drop it before piping to a resolver —
otherwise `dig` dutifully looks up a host named `NA`:

```sh
Rscript -e 'h <- punycoder::host_normalize(commandArgs(TRUE)); cat(h[!is.na(h)], sep = "\n")' münchen.de 'bad..domain' |
  xargs -n 1 dig +short
```

## Key Features

- **Reliable Encoding/Decoding**: RFC 3492 compliant punycode conversion
- **High Performance**: Vectorized operations for processing large datasets
- **Comprehensive Validation**: Robust error handling with informative messages
- **Flexible Backend**: Automatically uses `libidn2` when available, with a built-in fallback backend

## Use Cases

### Data Analysis
Clean and standardize URL datasets:

```{r eval=FALSE}
# Identify international domains
is_idn(c("café.com", "example.com", "москва.рф"))

# Validate domain names
validate_domain(c("valid.com", "invalid..domain"))
```

## Non-goals

`punycoder` is a standards primitive for Punycode and host normalization. It is
deliberately agnostic about resolvability and safety; the following are **not**
part of its acceptance criteria:

- **No spoof / homograph / mixed-script / display-safety detection.**
  `host_normalize()` is not a safety gate — a successful result says the host is
  valid and normalized under the pinned UTS #46 profile, nothing about whether it
  is visually safe or non-deceptive. Confusable and restriction-level checks
  (UTS #39 / UTR #36, which UTS #46 itself recommends only as application/UI-layer
  steps) belong upstack.
- **No URL parsing or canonicalization.** RFC 3986 / WHATWG URL parsing and
  canonicalization belong upstack; see the `rurl` package, which consumes
  punycoder's host functions.
- **No DNS resolvability or registrability / PSL classification.**
- **No address parsing.** There is no `email`-to-ASCII helper; splitting an
  address and IDNA-encoding its domain part is an addressing concern for an
  upstack consumer, not a Punycode primitive.
- **No per-TLD repertoire / allowed-character validation.** `host_normalize()`
  validates against the pinned UTS #46 profile, not against registry-specific
  IDN tables (which evolve independently of Unicode). TLD policy belongs upstack.

These opinions belong in higher layers that consume punycoder's host functions.

## How it compares

`punycoder` descends from the libidn-based
[`hrbrmstr/punycode`](https://github.com/hrbrmstr/punycode) and follows
IDNA2008 / UTS #46 rather than IDNA2003. The
[Prior art and comparison](https://bart-turczynski.gitlab.io/punycoder/articles/comparison.html)
article compares it with
Punycode libraries in other ecosystems and with the comparable R packages, on
the same inputs.

## Acknowledgments

These packages build on data, libraries, and prior work from many others.
See [ACKNOWLEDGMENTS.md](https://gitlab.com/bart-turczynski/punycoder/-/blob/main/ACKNOWLEDGMENTS.md) for the full list of thanks.

## Related packages

`punycoder` is part of a small ecosystem of R packages by the same author:

- **[pslr](https://CRAN.R-project.org/package=pslr)** — Public Suffix List engine that uses `punycoder` for IDNA canonicalization. Use it for eTLD and registrable-domain queries.
- **[rurl](https://CRAN.R-project.org/package=rurl)** — Full URL parsing, normalization, and joining toolkit built on top of both `punycoder` and `pslr`.

## Citation

If you use `punycoder` in your work, please cite it. Run `citation("punycoder")`
for the current citation, or see [`CITATION.cff`](https://gitlab.com/bart-turczynski/punycoder/-/blob/main/CITATION.cff).

Each release is archived on Zenodo. Cite the concept DOI
[10.5281/zenodo.20973629](https://doi.org/10.5281/zenodo.20973629) to refer to
the software in general (it always resolves to the latest version), or the
version-specific DOI shown on the [Zenodo
record](https://doi.org/10.5281/zenodo.20973629) for a particular release.

## Contributing

We welcome contributions. See [CONTRIBUTING.md](https://gitlab.com/bart-turczynski/punycoder/-/blob/main/CONTRIBUTING.md) for the
current development workflow,
[ARCHITECTURE.md](https://gitlab.com/bart-turczynski/punycoder/-/blob/main/ARCHITECTURE.md)
for how the package is structured, and
[DECISIONS.md](https://gitlab.com/bart-turczynski/punycoder/-/blob/main/DECISIONS.md)
for the design-decision log.

## Code of Conduct

Please note that this package is released with a [Contributor Code of Conduct](https://ropensci.org/code-of-conduct/). By contributing to this project, you agree to abide by its terms.

## License

MIT
