---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# punycoder

<!-- badges: start -->
[![R build status](https://github.com/bart-turczynski/punycoder/workflows/R-CMD-check/badge.svg)](https://github.com/bart-turczynski/punycoder/actions)
[![CRAN status](https://www.r-pkg.org/badges/version/punycoder)](https://CRAN.R-project.org/package=punycoder)
<!-- badges: end -->

High-performance Unicode and Punycode encoding/decoding for internationalized domain names (IDNs) in R.

## Overview

The `punycoder` package addresses critical gaps in R's URL processing capabilities by providing reliable, fast conversion between Unicode and ASCII representations of domain names. It follows RFC 3492 standards and is designed for robust handling of internationalized domain names in web scraping, data analysis, and URL processing workflows.

## Installation

You can install the development version of punycoder from
[GitHub](https://github.com/bart-turczynski/punycoder) with:

```{r eval=FALSE}
# install.packages("remotes")
remotes::install_github("bart-turczynski/punycoder")
```

### Optional native backend (`libidn2`)

`punycoder` works without extra system libraries. If `libidn2` is available at
build time, the package enables a native backend automatically; otherwise it
uses the built-in C++ fallback backend.

To install the recommended optional dependency:

- macOS (Homebrew):
  - `brew install libidn2 pkg-config`
- Debian/Ubuntu:
  - `sudo apt-get install libidn2-0-dev pkg-config`
- Fedora/RHEL/CentOS:
  - `sudo dnf install libidn2-devel pkgconf-pkg-config`
- Arch Linux:
  - `sudo pacman -S libidn2 pkgconf`

Verify the library is visible before installing `punycoder` from source:

```{r eval=FALSE}
system("pkg-config --modversion libidn2")
```

Then install/reinstall `punycoder`:

```{r eval=FALSE}
remotes::install_github("bart-turczynski/punycoder")
```

## Example

```{r example}
library(punycoder)

# Basic encoding
puny_encode("café.com")

# Check if domain is punycode
is_punycode("xn--example")

# Validate domains
validate_domain("test.com")
```

## Key Features

- **Reliable Encoding/Decoding**: RFC 3492 compliant punycode conversion
- **URL-Aware Processing**: Handle complete URLs with international domains  
- **High Performance**: Vectorized operations for processing large datasets
- **Comprehensive Validation**: Robust error handling with informative messages
- **CRAN Compliant**: Follows all CRAN policies and best practices

## Use Cases

### Web Scraping
Process international websites with Unicode domain names:

```{r eval=FALSE}
international_urls <- c(
  "https://café.paris.fr/menu",
  "https://москва.рф/news",
  "https://北京.中国/info"
)

# Convert for HTTP requests
ascii_urls <- url_encode(international_urls)
```

### Data Analysis
Clean and standardize URL datasets:

```{r eval=FALSE}
# Identify international domains
is_idn(c("café.com", "example.com", "москва.рф"))

# Validate domain names
validate_domain(c("valid.com", "invalid..domain"))
```

## Development Status

**Status**: Core punycode/IDN functionality is implemented and tested.

The package provides RFC 3492 encoding/decoding, URL host processing, and domain validation utilities.

## Contributing

We welcome contributions! Please see our [Contributing Guide](CONTRIBUTING.md) for details.

## License

GPL-3 
